Is high social class always beneficial for survival? A study of Northern Sweden

### 2017-11-22 article3.Rmd
Making last corrections, and adding a reference

---
title: Is high social class always beneficial for survival? A study of Northern Sweden
  1801--2013
author: "Sören Edvinsson and Göran Broström"
date: '`r Sys.Date()`'
output:
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
bibliography: mybib.bib
documentclass: article
fontsize: 12pt
citation_package: natbib
biblio-style: apalike
---

@@

Based on the results, we argue that high social class is not necessarily always favourable for survival. Social conditions and social position have impact, but not always in the expected direction. In our case this is apparent for men during a large part of the studied period; only at a surprisingly late date there appear a male mortality class reversal. Mortality risks in different contexts must be understood in the intersection between class and gender. We suggest that health-related behaviour is not only important in present-day societies, but was also decisive in earlier phases of the mortality transition. The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.

@@

A completely different pattern appears among men, which is the most surprising finding in
our analyses. The groups we expect to have the lowest mortality, if
we assume that access to vital resources and belonging to a high social class
determines survival, instead have by far the highest hazards of dying
during the 19th century and much of the 20th. Workers, on the
other hand, have comparatively low mortality until 1950, at the same
level as the middle class. The fact that we define social class as a time-dependent variable could otherwise lead to reverse causation -- those having bad health losing social position -- but instead, we find better survival among workers compared to the elite.

@@

Studies based on micro-data on adult and elderly populations in
history have been rare, mainly due to lack of adequate data. The results
from such studies show other patterns than that for infant and child
mortality. While some report substantial differences, many -- particularly those based on
Swedish data -- have found surprisingly small or even reversed socio-economic
differences during the 19th and early 20th centuries, sharply contrasting from the situation in present-day society. The expected social gradient in survival -- meaning that people
from higher social classes always had an advantage -- was in many cases,
particularly among adult men, not found despite the well-documented
socio-economic differences in society, for example in the Sundsvall region [@edli11; @edbr12]. This raises the question of when the modern social gradient appeared, if this was something that developed in the early parts of the 20th century. The results presented here show that the consistent advantage for higher social classes does not appear until surprisingly late in the investigated regions. The basic pattern is primarily observed in the working-age population, while the differences are smaller or even non-existent in old age in accordance with the status-levelling hypothesis.

The results showing a small or even reversed socio-economic gradient in the past and its late development are in accordance with other studies [@bevp11], for example @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, albeit with somewhat different social categories. They find the first indications of an emerging social gradient to be a fairly recent phenomenon, in their case as late as the 1950s. Similar results have also been found in Estonia [@jaadla17]. A study of mortality and social class in Sweden during the period 1960--1968 found that self-employed men and non-manual male workers in fact had higher mortality than manual workers in total as well as in cardiovascular mortality [@vano89]. A recent study Eriksson and Dribe on the social pattern of mortality, based on 1880---1910 census data linked to death registers shows that the unexpected pattern found in the present study is not unique, but rather seems to have characterised all parts of the country. These findings are contrary to what is usually assumed and what other studies have shown. The studies usually referred to when discussing the historically large social inequalities in health are mainly based on data from England and Wales [@ant67; @wowi95; @pamuk85]. We know less about how general these results are, but apparently the Swedish context differed from the English and Welsh ones.

@@

We believe our results have implications on several issues concerning our understanding of the demographic and epidemiologic development. One of the central theories has been presented by @mck76, who argues that improved nutrition has been the main determinant of the increase in life expectancy in recent centuries. This ought to imply that people with less adequate resources are the most vulnerable in the earlier phases of the mortality transition. While the McKeown thesis has been refuted by several scholars [@szreter88, @Colgrove02], it has played a central role within epidemiology and social medicine.

Our study seems thus to be in conflict with many of the dominating theories in present-day health research. In an overview of explanations of the persistence of health inequalities (primarily focusing on high-income countries with extensive welfare arrangements), @mack12 identifies a couple of theories, emphasizing different mechanisms and circumstances such as psycho-social pathways, material conditions, cultural capital, etc. However, most of these explanations are based on the assumption that higher social classes are always advantaged when it comes to health and mortality.

One of these leading theories is the hypothesis by @liph95 stating that social conditions are fundamental causes of disease and not only expressions of different disadvantages connected to social position. The criticism of the arguments by McKeown forced them to address following question:  "If Colgrove is right about the McKeown thesis, social epidemiology is left with a gaping hole in its explanatory repertoire and a challenge to a cherished principle about the importance of social factors of health" [@liph02]. This led Link and Phelan to incorporate alternative ways of how social conditions determine health, but still with the understanding that access to more resources guarantees better survival chances.Others emphasize psycho-social factors, for example @mar04, who argues that higher status leads to better health regardless of where on the status ladder the individual is situated. Psychosocial factors are also suggested by @wipi09 as an explanation for why equal societies perform better than unequal ones when it comes to health and other social conditions. In the long-term perspective provided here, neither of these theories fits the historical observations; their validity is mainly apparent for the most recent decades. To be fair, however, our study does not explicitly test for the possible impact of different material or psycho-social mechanisms. These factors could still be (and probably are) valid and influence the social pattern of mortality. What our study establishes is that there is no necessary eternal social gradient favouring higher social classes. Our results do however point to the important role of social position. The lacking social hierarchy may be caused by other socially determined mechanisms having a stronger impact on the pattern. We might have to consider the possibility of socially countervailing mechanisms [@lufr05].

@@

We must thus consider other and potentially more important explanations than the pure advantage of being in a high social position, reflecting that the embodiment can be materialized in different ways. In the following we discuss three possible pathways -- the life course perspective, social stress and class- and gender-based lifestyles. Concerning the life course perspective, exposure to diseases and other negative experiences in childhood can influence one'shealth later in life. Up until the early 20th century, being exposed to an urban environment had far higher health consequences than  living in rural Sweden. We control for the type of environment people live in, but we do not have full information about previous residences. There is a possibility that the elite group to a greater extent had spent their childhood in the more harmful urban environments, and that this might have influenced their health and survival in adulthood. It seems improbable, however, that differences in residential backgrounds could explain most of the observed pattern. The second pathway we discuss relates to the health risks of belonging to a certain social class. We usually assume that health risks are higher in social classes with less access to resources, as discussed earlier in this article. It is possible, however, that men in high positions had a stressful life, for example entrepreneurs facing the risk of losing an advantageous position. Business often failed which meant a substantial loss of prestige and social network. Suicides certainly occurred among the elite, but it is difficult to verify the extent of social stress in this and other social classes.

Social stress can also be expressed in unhealthy lifestyles, the third pathway we discuss. Harmful behaviours can be connected to certain expectationsassociated with being a man of high status. Our results regarding the gender aspect are important in this respect. As several researchers have pointed out, demographic research has often ignored power relations and different roles within families [@watkins93; @mack95; @jans07]. The gender difference in mortality may be caused by discrimination of the female party in the household. Historical Sweden was strongly gender-segregated, discriminating against women and giving them less space for action. In our study, we do not find the lack of power to have negative health consequences for women. Within the same households, the effect of social class leads to different results, high social position being beneficial for women but not for men. Instead, we might consider the role of behaviours based on distinct gendered and class expectations. Having a high social status is sometimes expressed in attitudes, norms and way of living that are potentially harmful to one's health. @edli11, finding similar results in another environment, suggest that an explanation for the larger social health inequalities in present-day society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking and smoking and being more inclined to take risks.  Alcohol has been suggested to be part of the large gender differences in the 19th century [@wil99; @edv92], and a more risk-taking lifestyle may be an important component of male mortality [@cour00]. Different historical contexts promote different ways to spend economic resources [@rasp06; @jaadla17].

In the today's high-income world including Sweden, smoking and drinking contribute to a substantial part of socio-economic health differences [@mapemy14; @syla16]. Our study suggests a strong impact in historical Sweden as well. If this interpretation is correct, the hypothesis by @ant67, on an increasing role of health-related behaviour in the modern low-mortality regime while economic inequality determined the social mortality pattern during the early phases of the mortality transition, can be questioned. Instead, health-related inequality might be equally strong earlier in history but expressed in a completely different way. To corroborate this claim, we need to find evidence on the social differentiation of consumption patterns and behaviours in history. Unfortunately, such data are rarely available before the 1960s. At best, we have circumstantial and anecdotal evidence. However, such evidence does support the idea of less health-beneficial lifestyles among elite men (@edli11 ).

@@

In this article we set our task as analysing social differences in mortality from a long-term perspective. Until recently it has not been possible to investigate the truly long-term development of mortality differences among adult and elderly populations. Still, the prevailing idea in many parts of the scholarly society is that strong socio-economic differences in health and mortality have always existed and were possibly even stronger in historical societies due to large class differences. The present study is unique in its time perspective, investigating the development from the early 19th century until the present, utilizing micro-data of high quality on a population of considerable size and social diversification. The resultswill likely surprise most readers; they partly surprised us as well. The consistent advantage of belonging to the elite has not always existed, and the modern pattern of social divide in health appeared much later than expected in the regions we studied. A male mortality class reversal has taken place. Apparently high social class is not always beneficial for survival.

However, there were large differences in the pattern between men and women as well as between different age groups. Class as well as gender must be considered. The social pattern of female mortality is in line with what was expected, but a social gradient in male adult mortality is a recent phenomenon. It is an interesting paradox hat both the class and gender aspects in the analysis show that those with more power and resources were in fact less successful when it came to survival. Furthermore, inequality in mortality appears clearer for the working-age population, while the differences are smaller among the elderly.

>>>>

---
title: Is high social class always beneficial for survival? A study of Northern Sweden
  1801--2013
author: "Sören Edvinsson and Göran Broström"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
bibliography: mybib.bib
documentclass: article
fontsize: 12pt
citation_package: natbib
biblio-style: apalike
---

@@

Based on the results, we argue that high social class is not necessarily always favourable for survival. Social conditions and social position certainly have impact, but not always in the expected direction. In our case this is apparent for men during a large part of the studied period; only at a surprisingly late date there appear a male mortality class reversal, changing the relation to a substantial advantage of being in a higher social position. Mortality risks in different contexts must be understood in the intersection between class and gender. We suggest that health-related behaviour is not only important in present-day societies, but was also decisive in earlier phases of the mortality transition. The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.

@@

A completely different pattern appears among men, which is the most surprising finding in
our analyses. The groups we expect to have the lowest mortality, if
we assume that access to vital resources and belonging to a high social class
determines survival, instead have by far the highest hazards of dying
during the 19th century and much of the 20th. Workers, on the
other hand, have comparatively low mortality until 1950, at the same
level as the middle category. The fact that we define social class as a time-dependent variable could otherwise lead to reverse causation -- those having bad health losing social position -- but instead, we find better survival among workers compared to the elite.

@@

Studies based on micro-data on adult and elderly populations in
history have been rare, mainly due to lack of adequate data. The results
from such studies show other patterns than that for infant and child
mortality. While some report substantial differences, many -- particularly those based on
Swedish data -- have found surprisingly small or even reversed socio-economic
differences during the 19th and early 20th centuries, sharply contrasting from the situation in present-day society. The expected social gradient in survival -- meaning that people
from higher social classes always had an advantage -- was in many cases,
particularly among adult men, not found despite the well-documented
socio-economic differences in society, for example in the Sundsvall region [@edli11; @edbr12]. This raises the question of when the modern social gradient appeared, if this was something that developed in the early parts of the 20th century. The results presented here show that the consistent advantage for higher social classes does not appear until surprisingly late in the investigated regions. Instead, the elite category experienced the highest death risks. The basic pattern is primarily observed in the working-age population, while the differences are smaller or even non-existent in old age in accordance with the status-levelling hypothesis.

The results showing a small or even reversed socio-economic gradient in the past and its late development are in accordance with other studies [@bevp11], for example @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, albeit with somewhat different social categories. They find the first indications of an emerging social gradient to be a fairly recent phenomenon, in their case as late as the 1950s. Compared to previous studies, the present one shows a much clearer survival disadvantage of the elite, and not only the lack of social gradient. There are in fact some studies indicating a similar pattern. A study of mortality and social class in Sweden during the period 1960--1968 found that self-employed men and non-manual male workers in fact had higher mortality than manual workers in total as well as in cardiovascular mortality [@vano89]. A recent study by @erdr17 on the social pattern of mortality, based on 1880--1910 census data linked to death registers, shows that the unexpected pattern found in the present study in not unique, but rather seems to have characterised all parts of the country. These findings are contrary to what is usually assumed and what other studies have shown. The studies usually referred to when discussing the historically large social inequalities in health are mainly based on data from England and Wales [@ant67; @wowi95; @pamuk85]. We know less about how general these results are, but apparently the Swedish context differed from the English and Welsh ones.

@@

We believe our results have implications on several issues concerning our understanding of the demographic and epidemiologic development. One of the central theories has been presented by @mck76, who argues that improved nutrition has been the main determinant of the increase in life expectancy in recent centuries. This ought to imply that people with less adequate resources are the most vulnerable in the earlier phases of the mortality transition. The results presented here prove that the most wealthy and privileged groups were not able to avoid the most common health risks in historical societies. This should make us cautious of accepting nutritional improvements as the most important explanation for the mortality transition. While the McKeown thesis has been refuted by several scholars [@szreter88, @Colgrove02], it has played and is still playing a central role within epidemiology and social medicine.

Our study seems thus to be in conflict with many of the dominating theories in present-day health research. In an overview of explanations of the persistence of health inequalities (primarily focusing on high-income countries with extensive welfare arrangements), @mack12 identifies a couple of theories, emphasizing different mechanisms and circumstances such as psycho-social pathways, material conditions, cultural capital, etc. However, most of these explanations are based on the assumption that higher social classes are always advantaged when it comes to health and mortality and that the different mechanisms all contribute to the advantage of higher social position.

One of these leading theories is the hypothesis by @liph95 stating that social conditions are fundamental causes of disease and not only expressions of different disadvantages connected to social position. The criticism of the arguments by McKeown forced them to address following question:  "If Colgrove is right about the McKeown thesis, social epidemiology is left with a gaping hole in its explanatory repertoire and a challenge to a cherished principle about the importance of social factors of health" [@liph02]. This led Link and Phelan to incorporate alternative ways of how social conditions determine health, but still with the understanding that access to more resources guarantees better survival chances.

Others emphasize psycho-social factors, for example @mar04, who argues that higher status leads to better health regardless of where on the status ladder the individual is situated. Psychosocial factors are also suggested by @wipi09 as an explanation for why equal societies perform better than unequal ones when it comes to health and other social conditions. In the long-term perspective provided here, neither of these theories fits the historical observations; their validity is mainly apparent for the most recent decades. To be fair, however, our study does not explicitly test for the possible impact of different material or psycho-social mechanisms. These factors could still be (and probably are) valid and influence the social pattern of mortality. What our study establishes is that there is no necessary eternal social gradient favouring higher social classes. Our results do however point to the important role of social position. The lacking social hierarchy may be caused by other socially determined mechanisms having a stronger impact on the pattern. We might have to consider the possibility of socially countervailing mechanisms [@lufr05].

@@

We must thus consider other and potentially more important explanations than the pure advantage of being in a high social position, reflecting that the embodiment can be materialized in different ways. In the following we discuss three possible pathways -- the life course perspective, social stress and class- and gender-based lifestyles. Concerning the life course perspective, exposure to diseases and other negative experiences in childhood can influence one's health later in life. Up until the early 20th century, being exposed to an urban environment had far higher health consequences than  living in rural Sweden. We control for the type of environment people live in, but we do not have full information about previous residences. There is a possibility that the elite group to a greater extent had spent their childhood in the more harmful urban environments, and that this might have influenced their health and survival in adulthood. It seems improbable, however, that differences in residential backgrounds could explain most of the observed pattern. The second pathway we discuss relates to the health risks of belonging to a certain social class. We usually assume that health risks are higher in social classes with less access to resources, as discussed earlier in this article. It is possible, however, that men in high positions had a stressful life, for example entrepreneurs facing the risk of losing an advantageous position. Business often failed which meant a substantial loss of prestige and social network. Suicides certainly occurred among the elite, but it is difficult to verify the extent of social stress in this and other social classes.

Social stress can also be expressed in unhealthy lifestyles, the third pathway we discuss. Harmful behaviours can be connected to certain expectations associated with being a man of high status. Our results regarding the gender aspect are important in this respect. As several researchers have pointed out, demographic research has often ignored power relations and different roles within families [@watkins93; @mack95; @jans07]. The gender difference in mortality may be caused by discrimination of the female party in the household. Historical Sweden was strongly gender-segregated, discriminating against women and giving them less space for action. In our study, we do not find the lack of power to have negative health consequences for women. Within the same households, the effect of social class leads to different results, high social position being beneficial for women but not for men. Instead, we might consider the role of behaviours based on distinct gendered and class expectations. Having a high social status is sometimes expressed in attitudes, norms and way of living that are potentially harmful to one's health. @edli11, finding similar results in another environment, suggest that an explanation for the large social health inequalities in present-day society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking and smoking and being more inclined to take risks.  Alcohol has been suggested to be part of the large gender differences in the 19th century [@wil99; @edv92], and a more risk-taking lifestyle may be an important component of male mortality [@cour00; @rasp06; @jaadla17]. Different historical contexts promote different ways to spend economic resources.

In the today's high-income world including Sweden, smoking and drinking contribute to a substantial part of socio-economic health differences [@mapemy14; @syla16]. Our study suggests a strong impact in historical Sweden as well. If this interpretation is correct, the hypothesis by @ant67, on an increasing role of health-related behaviour in the modern low-mortality regime while economic and material conditions determined the social mortality pattern during the early phases of the mortality transition, can be questioned. Instead, health-related inequality might be equally strong earlier in history but expressed in a completely different way. To corroborate this claim, we need to find evidence on the social differentiation of consumption patterns and behaviours in history. Unfortunately, such data are rarely available before the 1960s. At best, we have circumstantial and anecdotal evidence. However, such evidence does support the idea of less health-beneficial lifestyles among elite men (@edli11 ).

@@

In this article we set our task as analysing social differences in mortality from a long-term perspective. Until recently it has not been possible to investigate the truly long-term development of mortality differences among adult and elderly populations. Still, the prevailing idea in many parts of the scholarly society is that strong socio-economic differences in health and mortality have always existed and were possibly even stronger in historical societies due to large class differences. The present study is unique in its time perspective, investigating the development from the early 19th century until the present, utilizing micro-data of high quality on a population of considerable size and social diversification. The results will likely surprise most readers; they partly surprised us as well. The consistent advantage of belonging to the elite has not always existed, and the modern pattern of social divide in health appeared much later than expected in the regions we studied. A male mortality class reversal has taken place. Apparently high social class is not always beneficial for survival.

However, there were large differences in the pattern between men and women as well as between different age groups. Class as well as gender must be considered. The social pattern of female mortality is in line with what was expected, but a social gradient in male adult mortality is a recent phenomenon. It is an interesting paradox that both the class and gender aspects in the analysis show that those with more power and resources were in fact less successful when it came to survival. Furthermore, inequality in mortality appears clearer for the working-age population, while the differences are smaller among the elderly.
